A Practical Optional Type System for Clojure

### 2012-12-06 dissertation/intro.tex
Lyndon corrections + others

The Clojure programming language is a dynamically typed dialect of Lisp invented
by Hickey \footnote{\url{http://clojure.org/}}, designed to run on popular platforms.
It emphasises functional programming with immutable data structures
and provides direct interoperability with its host platform.
Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
the Common Language Runtime, and for Javascript virtual machines.
At the current time, Clojure on the JVM is the most mature implementation,
and therefore this project focuses on the JVM implementation.

@@

The initial motivation for implementing an optional type system
for Clojure was an anecdotal account of the struggles
of a Clojure programmer. In an apparently heroic effort, he managed to 
implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
an advanced form of ``pipes'', using arrows, a generalisation of monads.
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

He highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness of the library.
Without a static type system, it is a significant task
to verify such an implementation as correct due to heavy use
of higher-order functions.
Secondly, to aid him while writing the library.

Since then, the Clojure community has shown significant interest in this work.
I also developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community as a Clojure Google Summer of Code
project for 2012~\footnote{\url{http://www.google-melange.com/gsoc/project/google/gsoc2012/ambrosebs/10001}}.
I have also been invited to speak on this project at the Clojure Conj
2012 conference in November, the main international conference related to Clojure.

@@

In this dissertation we develop an optional type system for Clojure.
We focus on the features needed to make a practical type system that
existing Clojure programmers can use.
We also developed a prototype based on Typed Racket~\cite{Tob10}.

@@

  \item This prototype is extended to support several features that are not present in Typed Racket.
        We support filtering sequences on anonymous functions that convey only negative type information
        (eg. non-nil), capturing a common Clojure idiom.

  \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
        occurrences of Java's \emph{null}. We make \emph{null} directly expressible
        in Typed Clojure's type system as \lstinline|nil|. This makes it reasonable for it to be a
        static type error if \emph{null} is potentially misused, an improvement over Java's static type system
        where \emph{null} cannot be expressed.

  \item We show how a combination of intersection types and occurrence typing enables us to accurately type check
        the most common usages of Clojure's sequence abstraction. This was particularly subtle, as 
        many base operations in the sequence abstraction take and/or return \lstinline|nil|, and
        each case could imply different information about a sequence.

  \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types.
        These types are sufficient for accurately associating and dissociating known keyword keys,
        and accurately looking up values associated with keyword keys in heterogeneous map types that have
        that keyword key present.
        This is sufficient for type checking many common idioms involving maps.

  \item We identify the main future issues in typing Clojure code. Multimethod definitions are problematic
        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
        Records have subtle semantics for dissociating keys, like dissociating a base key of a record returns
        a plain map.
\end{itemize}

@@

(\lstinline|(defn name [...] body)| expands to \lstinline|(def name (fn [...] body))|,
where \lstinline|(def name init)| defines a new var in the current namespace 
called \lstinline|name|, and \lstinline|fn| creates a function value).

In this example, we define a new var \lstinline|collatz|
(when unambiguous, I omit the qualifying namespace/package for the remainder of the chapter).
Typed Clojure requires all 
used vars to be annotated. Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
annotates \lstinline|collatz| to be a function from \lstinline|Number| to 
\lstinline|Number| (\lstinline|Number| refers to \lstinline|java.lang.Number|,
due to every Clojure namespace implicitly importing all Classes in the \lstinline|java.lang| package,
the equivalent of Java's \lstinline|import java.lang.*;|).

>>>>

The Clojure programming language is a dynamically typed dialect of Lisp invented
by Hickey~\cite{Hic08}, designed to run on popular platforms.
It emphasises functional programming with immutable data structures
and provides direct interoperability with its host platform.
Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
the Common Language Runtime, and for Javascript virtual machines.
At the current time, Clojure on the JVM is the most mature implementation,
and therefore this project focuses on the JVM implementation.

@@

\subsubsection{Better error messages}

Dynamically typed languages like Clojure often sacrifice some of the advantages of
static typing. In particular, statically typed languages generally offer 
compile-time type errors where dynamically typed languages either give runtime
type errors, or none at all.

This is distinction is easily demonstrated by comparing Clojure and Haskell's behaviour at their interactive prompts.
Listing \ref{lst:cljrtime} shows how Clojure happily compiles a nonsensical function that performs addition
on a string. It is only until we invoke it that we get a type error.

\begin{lstlisting}[caption=Runtime type errors in Clojure, label=lst:cljrtime]
user=> (fn [x] (+ "string" x))
#<user$eval128376$fn__128377 user$eval128376$fn__128377@183ab4e>

user=> ((fn [x] (+ "string" x)) 1)
#<ClassCastException java.lang.ClassCastException: java.lang.String cannot be cast to java.lang.Number>
\end{lstlisting}

The error provided by Clojure is unsatisfying:

\begin{itemize}
  \item The error message does little to help the programmer identify the source of the error.
  \item It requires at least some knowledge of Clojure's internals to debug.
  \item We would rather receive the error at the earliest time possible: compile-time.
\end{itemize}

Listing \ref{lst:hskctime} shows how Haskell, a statically typed language, handles this interaction.
Haskell detects our programming error at compile-time, disallowing our function to compile.

\begin{lstlisting}[caption=Compile-time type errors in Haskell, label=lst:hskctime]
Prelude> (\x -> "string" + x)

<interactive>:1:17:
   No instance for (Num [Char])
   arising from a use of `+'
   Possible fix: add an instance declaration for (Num [Char])
   In the expression: "string" + x
   In the expression: (\ x -> "string" + x)
   In an equation for `it': it = (\ x -> "string" + x)
\end{lstlisting}

Haskell's error here is much preferred over Clojure's:

\begin{itemize}
  \item The exact source of the error is given in \emph{user code}.
  \item The error is given as early as possible: compile-time.
\end{itemize}

By adding static type checking to Clojure, we can achieve errors messages that are closer to
Haskell's.
Listing \ref{lst:tcctime} gives an idea of what kind of error messages Typed Clojure actually
provides in these situations.

\begin{lstlisting}[caption=Compile-time errors in Typed Clojure, label=lst:tcctime]
user=> (cf (fn> [[a :- Number]] (+ "string" a)))
#<Exception java.lang.Exception: 5: Static method clojure.lang.Numbers/add could not be applied to arguments:
Domains: 
  typed.core/AnyInteger typed.core/AnyInteger
  java.lang.Number java.lang.Number

Arguments:
  (Value "string") java.lang.Number

in: (. clojure.lang.Numbers add "string" a)>
\end{lstlisting}

Typed Clojure lists each of the expected domain types, of which the static method \lstinline|add| of the Java class
\lstinline|clojure.lang.Numbers| has been assigned two,
and also lists the types of the given arguments.
A line number and the source of the error is also provided.

Listing \ref{lst:tcctime} deserves some explanation:

\begin{itemize}
  \item The given source of the error may be surprising, but it is explained
        by noting that Clojure ``inlines'' \lstinline|(+ "string" a)| to a Java method call
        \lstinline|(. clojure.lang.Numbers add "string" a)|.
        \lstinline|cf| type checks a form.
  \item \lstinline|add|'s type has been overridden by Typed Clojure to be more specific (see Section \ref{sec:javainterop}).
  \item \lstinline|cf| type checks a form.
  \item \lstinline|(fn> [[a :- Number]] ...)| defines a function with its first argument of static type \lstinline|Number|.
\end{itemize}

\subsubsection{Higher order programming}

The initial motivation for implementing an optional type system
for Clojure was an anecdotal account of the struggles
of a Clojure programmer, Jim Duey. In an apparently heroic effort, Duey managed to 
implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
an advanced form of ``pipes'', using arrows, a generalisation of monads.
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

Duey highlighted a strong desire for a type system for several reasons.
Firstly, to verify the correctness of the library.
Without a static type system, it is a significant task
to verify such an implementation as correct due to heavy use
of higher-order functions.
Secondly, to aid him while writing the library.

Since then, the Clojure community has shown significant interest in this work.
I also developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community as a Clojure Google Summer of Code
project for 2012~\cite{GSOC2012}.
I have also been invited to speak on this project at the Clojure Conj
2012 conference in November, the main international conference related to Clojure.

@@

In this dissertation we develop an optional type system for Clojure.
We focus on the features needed to make a practical type system that
existing Clojure programmers can use.
We also develop a prototype based on Typed Racket~\cite{Tob10}.

@@

  \item This prototype is extended to support several features that are not yet present in Typed Racket.
        We support filtering sequences on anonymous functions that convey only negative type information
        (eg. non-nil), capturing a common Clojure idiom.
        We enable type checking of very abstract code (like code using monads)
        by adding \emph{type functions} (functions at the type level).

  \item We support safer interoperability with Java. We show how techniques developed for
        Typed Racket, such as occurrence typing~\cite{TF10},  of Java's \emph{null}
        while still conforming to existing Clojure idioms.

  \item We can type check several particularly useful Clojure idioms, like common usages of the \emph{sequence abstraction}
        (Clojure's common interface to all collections) and 
        common usages of ``untyped'' hash-maps, which have a similar role to records in ML or Haskell,
        and objects in object-oriented languages.

  \item We identify the main future issues in typing Clojure code. Multimethod definitions are problematic
        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
        Records have subtle semantics for dissociating keys, specifically: dissociating all base keys of a record returns
        a plain map.
\end{itemize}

@@

(\lstinline|(defn collatz [n] ...)| expands to \lstinline|(def collatz (fn [n] ...))|,
where \lstinline|(def name init)| defines a new var in the current namespace 
called \lstinline|name|, and \lstinline|fn| creates a function value).

In this example, we define a new var \lstinline|collatz|
(when unambiguous, I omit the qualifying namespace/package for the remainder of the chapter).
Typed Clojure type checks at the namespace granularity. All vars that occur
in a ``typed'' namespace must have a type annotation. 
Here \lstinline|typed.core/ann|, a Typed Clojure macro for annotating vars, 
annotates \lstinline|collatz| to be a function from \lstinline|Number| to 
\lstinline|Number| (\lstinline|Number| refers to \lstinline|java.lang.Number|,
due to every Clojure namespace implicitly importing all Classes in the \lstinline|java.lang| package,
the equivalent of Java's \lstinline|import java.lang.*;|).

### 2012-12-06 dissertation/lit-review.tex
Lyndon corrections + others

Scheme then moved to program analysis techniques like Set Based Analysis (SBA).
They, however, suffered from modularity issues due to the lack of declared intended types.
MrSpidey\footnote{\url{http://www.plt-scheme.org/software/mrspidey/}}
is a static type checker for Racket that uses uses SBA.
Felleisen states the main problem with these kinds of systems: ``Mr Spidey and friends infer brittle and large
types; errors remain difficult to explain and fix''~\cite{Fell09}.

@@

Gradual typing combines static and runtime type checking, so programmers
can utilise whichever is most appropriate for a given situation.
A key feature of gradual type systems is its use of \emph{bidirectional checking}
like~\cite{PT00}, which requires type annotations in natural places like functions.
Type errors in gradual type systems are often more manageable than in Soft Typing and SBA,
often more comparable to type errors in statically typed languages.

Gradually typed languages have different degrees of runtime checking.
Typed Racket was developed as a path for module-by-module
porting of existing untyped Racket modules to a typed sister language~\cite{Tob10}.
Once a module is ported and type checked, it is protected from untyped modules
by inserting runtime checks.
Typescript~\cite{TypeS2012}, a gradual type system for Javascript, 
does not add runtime checks to untyped interactions.
Runtime checks affect performance, but give better errors and ensure
type errors result in the program failing quicker.

@@

Clojure is a dynamic functional language hosted on the Java Virtual Machine. It provides 
interoperability with Java libraries. As Clojure is a dynamically typed language, it does
not give strong type type guarantees at compile time whether interactions with Java
are type safe.

@@

Typed Racket includes safe interoperability between any combination of typed and untyped 
Racket modules~\cite{Tob10,TF08}. 
Interactions with untyped modules are protected by adding runtime checks based on expected types.
Typed Racket implements a sophisticated blame calculus, based on Wadler~\cite{WF09}. It ensures 
error messages correctly \emph{blame} the source of type errors,
which can be difficult to determine in the presence of higher-order functions.

@@

O'Caml-style extensible record types has been the subject of extensive research 
(eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91}).
Typed Clojure's heterogeneous map types show some resemblance to extensible
record types, but a survey of the literature did not reveal research flexible
enough to capture common usages of maps in Clojure.
This most likely implies that designing such a system for Typed Clojure would be hard.
For this reason, heterogeneous map types are kept as simple as possible while still being useful.

@@

Other statically typed languages have similar goals
to support Java interoperability, like, most completely, Scala.
Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
with Java's \emph{null} pointer by using an \emph{Option} type, which (if
used correctly) provides strong elimination for \emph{null}.
\emph{null} is still not expressible as a type in Scala, however.
Arrays are parameterised by a non-variant (or invariant) parameter,
which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
this checks that covariant parameters are only used in covariant positions.

@@

In untyped languages like Scheme, functions can take any number of arguments. It is an
error to supply less than the minimum number of arguments to a function.
This style features sophisticated support for functions with variable-arity. For example,
functions can dispatch on the number of arguments provided, and supports a \emph{rest} parameter
as its last parameter which can accept any number of arguments.

@@

Strickland et al.\ invented a type system supporting variable-arity polymorphism ~\cite{STF09}
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

For example in Clojure, the function \lstinline|map| takes a function and one or more sequences,
and returns the result of applying the function argument to each element of the sequences pair-wise.

@@

The pseudocode in listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
to infer the type arguments to \lstinline|map| by observing the argument types of the first
argument to \lstinline|map| and the types of subsequent sequence arguments. Bidirectional checking
then \emph{synthesises} the resulting type of the expression by \emph{checking} each element
of \lstinline|[1 2]| is a subtype of \lstinline|a|, and each element of \lstinline|[2.1 3.2]| is a subtype of
\lstinline|b|. The result of the anonymous function argument is \emph{synthesised} from the type of
\lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
synthesise the type of listing \ref{lst:bidir} to be \lstinline|(LazySeq Float)|.

@@

Dynamically typed languages use an ad-hoc combination of type predicates,
selectors, and conditionals to steer execution flow and reason about runtime types of variables.
Typed Racket uses occurrence typing to capture these ad-hoc type refinements.
For example, listing \ref{lst:occ1} shows occurrence typing following the implications 
of the type predicate \emph{number?} and the selector \emph{first}, and utilises those implications to refine
the type of \lstinline|x|. If the test at line 3 succeeds, occurrence typing refines the
type of \lstinline|(first x)| to be \lstinline|Number|, which allows \lstinline|(+ 1 (first x))|
to be well typed. Similarly at line 4, we can be sure that \lstinline|(first x)| is
a \lstinline|String|, since we have ruled out the case of being a \lstinline|Number|.

@@

Occurrence typing~\cite{TF08,TF10} extends the type 
system with a \emph{proposition environment} that represents 
the information on the types of bindings down conditional branches.
These propositions are then used to update the types associated
with bindings in the \emph{type environment} down branches
so binding occurrences are given different types 
depending on the branches they appear in, and the conditionals
that lead to that branch.
(See section \ref{sec:occurenceimpl} for details on how Typed Clojure uses occurrence typing).

@@

An exciting application of occurrence typing as yet unexplored is facilitating null-safe interoperability with Java.
By declaring \lstinline|nil| (Clojure's value of Java's \lstinline|null|) to \emph{not} be a subtype of reference types,
we can begin to statically disallow potentially inconsistent usages of \lstinline|nil| as part of the type system.

@@

Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()| that
that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
\lstinline|Object| static type, as \lstinline|null| is a member of all reference types. By surrounding
the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
\lstinline|a| is non-nil when passed to \lstinline|expects-non-nil|. Occurrence typing infers
this by observing \lstinline|nil| is a false value in Clojure, therefore \lstinline|a| cannot
be \lstinline|nil| the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
from \lstinline|(U nil Object)|.

@@

After some investigation, typing multimethods with Typed Clojure is assigned as future work
(see section \ref{sec:mmfuture}).

@@

Many advanced type systems provide support for \emph{higher kinds},
which are ``another level up'' from types.
For example, a type \lstinline|Number| is distinguished from a
\emph{type constructor} \lstinline|Monad| (see section \ref{sec:monads}), which is a type level function
(similar to the difference between values and functions on the value level, but on the type level).

F$_\omega$ is a typed $\lambda$-calculus with support for higher kinds,
specifically type constructors (see Pierce~\cite{Pie02}).
Haskell~\cite{Mar10} distinguishes between types and type constructors,
the latter of which is an essential part of its monad library 
(see section \ref{sec:monads} for a monad library ported to Typed Clojure
that uses type constructors similarly).

>>>>

Scheme then moved to program analysis techniques like Set-Based Analysis~\cite{FF97} (SBA).
They, however, suffered from modularity issues due to the lack of declared intended types.
MrSpidey\footnote{\url{http://www.plt-scheme.org/software/mrspidey/}}
is a static type checker for Racket that uses uses SBA.
Felleisen states the main problem with these kinds of systems: ``Mr Spidey and friends infer brittle and large
types; errors remain difficult to explain and fix''~\cite{Fell09}.

@@

Gradual typing combines static and runtime type checking, so programmers
can choose the most appropriate one for the situation.
A key feature of gradual type systems is its use of \emph{bidirectional checking}
like~\cite{PT00}, which requires type annotations in natural places like functions.
Type errors in gradual type systems are often more manageable than in Soft Typing and SBA,
often more comparable to type errors in statically typed languages.

Gradually typed languages have different degrees of runtime checking.
Typed Racket was developed as a path for module-by-module
porting of existing untyped Racket modules to a typed sister language~\cite{Tob10}.
Once a module is ported and type checked, it is protected from untyped modules
by inserting runtime checks.
Typescript~\cite{TypeS2012}, a gradual type system for Javascript, 
does not add runtime checks to untyped interactions.
Runtime checks affect performance, but give better errors and ensure
type errors result in the program failing more quickly.

@@

Clojure is a dynamic functional language hosted on the Java Virtual Machine. It provides 
interoperability with Java libraries. As Clojure is a dynamically typed language, it does
not give strong type guarantees at compile time that interactions with Java
are type safe.

@@

Typed Racket includes safe interoperability between any combination of typed and untyped 
Racket modules~\cite{Tob10,TF08}. 
Interactions with untyped modules are protected by adding runtime checks based on expected types.
Typed Racket implements a sophisticated blame calculus, based on Wadler~\cite{WF09}. It ensures 
error messages always correctly \emph{blame} the source of type errors,
which can be difficult to determine in the presence of higher-order functions.

@@

O'Caml-style extensible record types have been the subject of extensive research 
(eg. Wand~\cite{Wan89}, Cardelli and Mitchell~\cite{CM91}, Harper and Pierce~\cite{HP91}).
Typed Clojure's heterogeneous map types show some resemblance to extensible
record types, but a survey of the literature did not reveal research flexible
enough to capture common usages of maps in Clojure.
This most likely implies that designing such a system for Typed Clojure would be hard.
For this reason, heterogeneous map types are kept as simple as possible while still being useful.

@@

Other statically typed languages have similar goals
to support Java interoperability, like, most completely, Scala.
Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
with Java's \emph{null} pointer by using an \emph{Option} type, which (if
used correctly) provides strong elimination for \emph{null}.
\emph{null} is still not expressible as a type in Scala, however.
Arrays are parameterised by a non-variant (or invariant) parameter,
which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
this checks that covariant parameters are used only in covariant positions.

@@

In untyped languages like Scheme, functions can have more complicated arity.
In this context, functions have a fixed number of required parameters, and an optional \emph{rest}
parameter, which represents the sequence of the arguments provided after the required parameters.
It is an error to apply a function to fewer arguments than the functions number of required parameters.
This style features sophisticated support for functions with variable-arity. For example,
Scheme functions can dispatch on the number of arguments provided, and support an optional \emph{rest} parameter
which can accept any number of arguments.

@@

Strickland \emph{et al.} invented a type system supporting variable-arity polymorphism ~\cite{STF09} % doesn't fit on a line
a version of which is included in the current implementation of Typed Racket.
Their main innovation centres around \emph{dotted type variables}, which represent a heterogeneous sequence
of types. Dotted type variables allow \emph{non-uniform} variable-arity function types,
which are used to check definitions and usages of functions with non-trivial rest parameters

For example in Clojure, the function \lstinline|map| takes a function and at least one sequence,
and returns a sequence of the results of applying the function to each subsequent element of the sequences simultaneously.

@@

The pseudocode in Listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
to infer the type arguments to \lstinline|map| by observing the argument types of the first
argument to \lstinline|map| and the types of subsequent sequence arguments. Bidirectional checking
then \emph{synthesises} the resulting type of the expression by \emph{checking} each element
of \lstinline|[1 2]| is a subtype of \lstinline|a|, and each element of \lstinline|[2.1 3.2]| is a subtype of
\lstinline|b|. The result of the anonymous function argument is \emph{synthesised} from the type of
\lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
synthesise the type of Listing \ref{lst:bidir} to be \lstinline|(LazySeq Float)|.

@@

Dynamically typed languages use an ad-hoc combination of type predicates,
selectors, and conditionals to steer execution flow and reason about runtime types of variables.
Typed Racket uses occurrence typing to capture these ad-hoc type refinements.
For example, Listing \ref{lst:occ1} shows occurrence typing following the implications 
of the type predicate \emph{number?} and the selector \emph{first}, and utilises those implications to refine
the type of \lstinline|x|. If the test at line 3 succeeds, occurrence typing refines the
type of \lstinline|(first x)| to be \lstinline|Number|, which allows \lstinline|(+ 1 (first x))|
to be well typed. Similarly at line 4, we can be sure that \lstinline|(first x)| is
a \lstinline|String|, since we have ruled out the case of being a \lstinline|Number|.

@@

Occurrence typing~\cite{TF08,TF10} extends the type 
system with a \emph{proposition environment} that represents 
the information on the types of bindings down conditional branches.
These propositions are then used to update the types associated
with bindings in the \emph{type environment} down branches
so binding occurrences are given different types 
depending on the branches they appear in, and the conditionals
that lead to that branch.
(See Section \ref{sec:occurenceimpl} for details on how Typed Clojure uses occurrence typing).

@@

An exciting application of occurrence typing as yet unexplored is facilitating null-safe interoperability with Java.
By declaring \lstinline|nil| (Clojure's value of Java's \lstinline|null|) to \emph{not} be a subtype of reference types,
we can begin to statically disallow potentially inconsistent usages of \lstinline|nil| as part of the type system.
\\\\
% fit next listing onto one page

@@

Listing \ref{lst:nil} infers from the Java signature \lstinline|Object getObject()|
that \lstinline|a| is of type \lstinline|(U nil Object)|. This is equivalent to Java's
\lstinline|Object| static type, as \lstinline|null| is a member of all reference types. By surrounding
the call \lstinline|(expects-non-nil a)| with \lstinline|(when a ...)|, we guarantee that
\lstinline|a| is non-nil when passed to \lstinline|expects-non-nil|. Occurrence typing infers
this by observing \lstinline|nil| is a false value in Clojure, therefore \lstinline|a| cannot
be \lstinline|nil| in the body of the \lstinline|when|, refining \lstinline|a|'s type to \lstinline|Object|
from \lstinline|(U nil Object)|.

@@

After some investigation, typing multimethods with Typed Clojure is assigned as future work
(see Section \ref{sec:mmfuture}).

@@

Many advanced type systems provide support for \emph{higher kinds},
which are ``another level up'' from types.
For example, a type \lstinline|Number| is distinguished from a
\emph{type constructor} \lstinline|Monad| (see Section \ref{sec:monads}), which is a type level function
(similar to the difference between values and functions on the value level, but on the type level).

F$_\omega$ is a typed $\lambda$-calculus with support for higher kinds,
specifically type constructors (see Pierce~\cite{Pie02}).
Haskell~\cite{Mar10} distinguishes between types and type constructors,
the latter of which is an essential part of its monad library 
(see Section \ref{sec:monads} for a monad library ported to Typed Clojure
that uses type constructors similarly).

### 2012-12-06 dissertation/paper.tex
Lyndon corrections + others

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions accepting an even number of variable arguments
are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.
Overall, the work described in this dissertation leads to the conclusion that it appears to 
be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

@@

Last November I attended my first conference, Clojure Conj 2011 (Raleigh, NC),
where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
A week later my supervisor Rowan Davies invited me to visit Carnegie Mellon University
where he was working on his PhD project. Most memorably, I was invited to a meeting between
Rowan and his PhD supervisor Frank Pfenning, which was a great thrill.

>>>>

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions with particular constraints on the number of 
variable arguments they accept (particularly, functions taking only an even number of variable arguments) are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.
Overall, the work described in this dissertation leads to the conclusion that it appears to 
be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

@@

Last November I attended my first conference, Clojure Conj 2011 (Raleigh, NC),
where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
A week later my supervisor Rowan Davies invited me to visit Carnegie Mellon University
where he worked on his PhD project. Most memorably, I was invited to a meeting between
Rowan and his PhD supervisor Frank Pfenning.
